---
title: "Activity_1 - Data Transformation"
output: html_notebook
---

```{r setup, include=FALSE}
library(readr)    # for loading data in a clean way
library(ggplot2)  # for making nice plots
library(cowplot)  # for making nicer plots
library(emmeans)  # for extracting the results of an analysis
library(lme4)     # for fitting models to more complex experimental designs
library(lmerTest) # for better hypothesis tests
library(pbkrtest) # for better hypothesis tests
library(multcomp) # for better hypothesis tests
library(multcompView) # for better hypothesis tests
library(PLS205)   # additional helper functions I've written for teaching
```

------------------------------------------------------------------------

# Background

## 1. Re-load the dataset if needed

```{r, results = 'hide', message=FALSE}
bolting_data_1 = read_csv('data/Bolting_experiment_1.csv')
bolting_data_1$mutant = factor(bolting_data_1$mutant,levels = c('WT','flc-3','vin3-4'))
```

## 2. Repeating plot1 of data

```{r}
plot1 = ggplot(bolting_data_1,aes(x=FRI,y=Days.to.Bolt)) + 
  geom_jitter(width = 0.1,aes(color = FRI)) + 
  facet_grid(~mutant) +
  expand_limits(y=0) + 
  theme_bw() + theme(legend.position = 'none') 
plot1
```

## Make the same plot with the data transformation

```{r}
plot2 = ggplot(bolting_data_1,aes(x=FRI,y=inverse(Days.to.Bolt/100))) + 
  geom_jitter(width = 0.1,aes(color = FRI)) + 
  facet_grid(~mutant) +
  expand_limits(y=0) + 
  theme_bw() + theme(legend.position = 'none') 
plot2
```

## Put the plots together to directly compare:

```{r}
plot_grid(plot1,plot2,ncol=1)
```

------------------------------------------------------------------------

# Activity 1

-   Re-do the analysis from the tutorial, this time using the `inverse(Days.to.bolt)` transformation.
-   Compare the results

## Original code

Here is the code from the tutorial:

```{r}
bolting_data_1 = read_csv('data/Bolting_experiment_1.csv')
bolting_data_1$mutant = factor(bolting_data_1$mutant,levels = c('WT','flc-3','vin3-4'))
str(bolting_data_1)

bolting_model = lm(Days.to.Bolt ~ FRI + mutant + FRI:mutant, bolting_data_1)

bolting_means = emmeans(bolting_model,specs = 'FRI',by = 'mutant')
print(bolting_means)

FRI_effects = contrast(bolting_means,'trt.vs.ctrl',ref = 'fri',name = 'FRI_effect')
print(FRI_effects)

regrouped_FRI_effects = update(FRI_effects,by = 'FRI_effect')
print(regrouped_FRI_effects)

mutant_effect_on_FRI_effects = contrast(regrouped_FRI_effects,'trt.vs.ctrl',ref = 'WT',name = 'mutant_effect')
print(mutant_effect_on_FRI_effects)

plot(FRI_effects,horizontal = FALSE) + 
  facet_grid(~mutant) +
  theme_bw()
```

## New analysis

Copy and modify the code here

**Hint**: If you save the plot of the FRI effects above in one variable, and then make a new plot and save it in a new variable, you'll be able to combine the two plots into a single figure to directly compare.

```{r}

```

------------------------------------------------------------------------

# Activity 2

Now, let's load the data from a different environment.

The file `data/Bolting_experiment_2.csv` contains data from the same experiment, but with the plants vernalized for 8 weeks.

-   Repeat the analysis (using the same data transformation)
-   Compare the results **visually**
-   After, we'll work through how to measure the effect of vernalization on the effect of FRI (for each mutant)!

```{r}

```

> **Questions**
>
> 1.  What mutant effects changed the most between the Vernalized (this experiment) and unvernalized (old experiment) conditions?
> 2.  What would you conclude about the effect of FRI on flowering time?
