---
title: "3-way factorial analysis"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)  # for making nice plots
library(cowplot)  # for making nicer plots
library(emmeans)  # for extracting the results of an analysis
library(lme4)     # for fitting models to more complex experimental designs
library(lmerTest) # for better hypothesis tests
library(pbkrtest) # for better hypothesis tests
library(multcomp) # for better hypothesis tests
library(multcompView) # for better hypothesis tests
library(PLS205)   # additional helper functions I've written for teaching
```

## Description of the experiment and goals

Burghardt, et al. (2016) repeated the same experiment in 16 conditions. 
Combinations of 2 temperatures, 2 diurnal patterns, 2 daylengths, and 2 vernalizations.
We'll focus on the 4 "warm", "long day" conditions:

- 22ConLDNV
- 22ConLDV
- 22VarLDNV
- 22VarLDV

We want to measure the effect of these environmental conditions on the FRI effects we studied in the first tutorial and activities.

## Load the full dataset
### Load the full dataset

```{r load_data,exercise=T}
bolting_data = read.csv('data/Bolting_experiment_full.csv')
```

### View the data

```{r view1,exercise=T, exercise.setup = 'load_data'}
str(bolting_data)
```

### Change variables to **factors**

```{r set_factors2,exercise=F, include = FALSE, exercise.setup = 'load_data'}
bolting_data$mutant = factor(bolting_data$mutant,levels = c('WT','flc-3','vin3-4'))
bolting_data$Genotype = factor(bolting_data$Genotype)
bolting_data$FRI = factor(bolting_data$FRI)
bolting_data$Treatment.V = factor(bolting_data$Treatment.V)
bolting_data$Chamber = factor(bolting_data$Chamber)
```


For complex datasets, it's helpful to use *R's* *factor* feature to label data as categorical.
Generally, most variables in an experiment should be treated as categorical, except the response.

To convert a variable to categorical, use the `as.factor()` function.

Note: we used this function before with the `mutant` variable do tell *R* which order we wanted. If we don't give an order, it will do it alphabetically.

Change all the categorical variables to factors below:

```{r set_factors,exercise=T, exercise.setup = 'load_data'}
# here is the first:
bolting_data$mutant = factor(bolting_data$mutant,levels = c('WT','flc-3','vin3-4'))
```

Then use `str(bolting_data)` to see the improved view:
```{r view2,exercise=T, exercise.setup = 'set_factors2'}
str(bolting_data)
```


## Run the analysis

The basic outline of the analysis will be the same as before, but now with another layer of regrouping and contrasting:

1.  Fit a model to the data
2.  Compute the cell means for each level of the **focal treatment** separately for each level of all **moderator treatments**.
3.  Compute the **focal treatment effects** separately for each level of all **moderator treatments**.
4.  Compare the focal treatment effects **among levels of** the **first** moderator treatments
5.  Compare the first moderator treatment effects on the focal treatment **among levels of** the **second** moderator treatment


### Fit the model

This is code to fit the model.

```{r model1,exercise=TRUE, exercise.setup = 'set_factors2'}
str(bolting_data)
bolting_model = lm(inverse(Days.to.Bolt) ~ FRI + mutant + Treatment.V + 
                                          FRI:mutant + FRI:Treatment.V + mutant:Treatment.V +
                                          FRI:mutant:Treatment.V, 
                                          data = bolting_data)
```

Note how we include all 3 treatment variables, then all 2-way combinations, then the 3-way combination!

There is a short-hand for this: `FRI*mutant*Treatment.V`. 
It's fine to use once you're comfortable with *R's* models, but be sure you understand the model statements first.


### Compute the treatment effects

Using the fitted model, we will compute the treatment effects of interest, using the following steps:

1.  Compute the cell means for each treatment combination
2.  Compute the **focal treatment** effects at each level of the moderators
3.  Regroup the effects by focal treatment effect
4.  Compute the effects of each moderator-1 level on each focal treatment, **grouped by level of moderator 2**
5. Regroup by levels of moderator-1
6. Compute the effects of changing moderator-2

### Compute the cell means for each treatment combination

```{r cell_means,exercise = TRUE,exercise.setup ='model1'}
bolting_means = emmeans(bolting_model,specs = 'FRI',by = c('mutant','Treatment.V'))
print(bolting_means)
```

### Compute the **focal treatment** effects at each level of the moderator

Now, here's how to do it with *R*:

```{r FRI_effects,exercise = TRUE,exercise.setup ='cell_means'}
FRI_effects = contrast(bolting_means,'trt.vs.ctrl',ref = 'fri',name = 'FRI_effect')
print(FRI_effects)
```


### Plot the FRI effect estimates

We can make a plot of all FRI effects, though it's a bit hard to read:
```{r plot_FRI_effects,exercise = TRUE,exercise.setup ='FRI_effects'}
plot(FRI_effects,horizontal = FALSE) + facet_grid(mutant~Treatment.V)
```

> **Activity** 
> 
> - Try flipping the two moderators in `facet_grid`. Do you see effects of both mutant and Treatment.V on the FRI effects?

### Compare the FRI effects among the mutants - Regroup

To compare effects, they have to be in the same `by`-group.
We need to regroup the FRI effects so they can be compared.

We want to do this separately for each environment (`Treatment.V`)

```{r regroup_FRI_effects,exercise = TRUE,exercise.setup ='FRI_effects'}
regrouped_FRI_effects = update(FRI_effects,by = c('Treatment.V','FRI_effect'))
print(regrouped_FRI_effects)
```

### Compare the FRI effects among the mutants

Now, compare the FRI effects for each mutant against the FRI effects for Col (*i.e.* `WT`) using `contrast():`

```{r mutant_FRI_effects,exercise = TRUE,exercise.setup ='regroup_FRI_effects'}
mutant_effect_on_FRI_effects = contrast(regrouped_FRI_effects,'trt.vs.ctrl',ref = 'WT',name = 'mutant_effect')
print(mutant_effect_on_FRI_effects)
```

This is basically repeating what you did in the second activity, but now repeating it 4 times, once for each of the four `Treatment.V` environments.

> Can you see changes in the effect of a mutant on the FRI effect among environments?

### Regroup the mutant effects on the FRI effect to enable comparisons

```{r regroup_mutant_FRI_effects,exercise = TRUE,exercise.setup ='mutant_FRI_effects'}
regrouped_mutant_effect_on_FRI_effects = update(mutant_effect_on_FRI_effects,by = c('mutant_effect','FRI_effect'))
print(regrouped_mutant_effect_on_FRI_effects)
```

### Compare the mutant effects on the FRI effect among the environments

We could choose a reference environment, or just compare all pairwise

```{r env_effects_mutant_FRI_effects,exercise = TRUE,exercise.setup ='regroup_mutant_FRI_effects'}
env_effects_on_mutant_effect_on_FRI_effects = contrast(regrouped_mutant_effect_on_FRI_effects,'pairwise')
print(env_effects_on_mutant_effect_on_FRI_effects)
```


> **Activity** 
> 
> - Come up with a way to summarize these results!
> - Hint: A plot might help!

```{r plot_env_effects_mutant_FRI_effects,exercise = TRUE,exercise.setup ='regroup_mutant_FRI_effects'}

```


### Alternative view using Compact Letter Display:

```{r cld_env_effects_mutant_FRI_effects,exercise = TRUE,exercise.setup ='regroup_mutant_FRI_effects'}
print(cld(regrouped_mutant_effect_on_FRI_effects,Letters = letters))
```

Note: for compact letter displays, ignore the "SE", "df", "t.ratio" and "p.value" columns!

## Full code:
```{r full_code, exercise = TRUE,exercise.lines = 39}
bolting_data = read.csv('data/Bolting_experiment_full.csv')

bolting_data$mutant = factor(bolting_data$mutant,levels = c('WT','flc-3','vin3-4'))
bolting_data$Genotype = factor(bolting_data$Genotype)
bolting_data$FRI = factor(bolting_data$FRI)
bolting_data$Treatment.V = factor(bolting_data$Treatment.V)

bolting_model = lm(inverse(Days.to.Bolt) ~ FRI + mutant + Treatment.V + 
                                          FRI:mutant + FRI:Treatment.V + mutant:Treatment.V +
                                          FRI:mutant:Treatment.V, 
                                          data = bolting_data)

bolting_means = emmeans(bolting_model,specs = 'FRI',by = c('mutant','Treatment.V'))
print(bolting_means)

FRI_effects = contrast(bolting_means,'trt.vs.ctrl',ref = 'fri',name = 'FRI_effect')
print(FRI_effects)

regrouped_FRI_effects = update(FRI_effects,by = c('Treatment.V','FRI_effect'))
print(regrouped_FRI_effects)

mutant_effect_on_FRI_effects = contrast(regrouped_FRI_effects,'trt.vs.ctrl',ref = 'WT',name = 'mutant_effect')
print(mutant_effect_on_FRI_effects)

regrouped_mutant_effect_on_FRI_effects = update(mutant_effect_on_FRI_effects,by = c('FRI_effect','mutant_effect'))
print(regrouped_mutant_effect_on_FRI_effects)

env_effects_on_mutant_effect_on_FRI_effects = contrast(regrouped_mutant_effect_on_FRI_effects,'pairwise')
print(env_effects_on_mutant_effect_on_FRI_effects)

print(cld(regrouped_mutant_effect_on_FRI_effects,Letters = letters))

plot(regrouped_mutant_effect_on_FRI_effects,horizontal = T) +
  geom_vline(xintercept = 0) +
  theme_bw()
```

